# Simulation^[Module originally written by Connor Jerzak and Shiro Kuriwaki] {#simulation}


```{r, include=FALSE}
library(dplyr)
library(purrr)

set.seed(7242) #set from Random.org
```

### Motivation: Simulation as an Analytical Tool {-}

An increasing amount of political science contributions now include a simulation. 

* [Axelrod (1977)](http://www-personal.umich.edu/~axe/research/Dissemination.pdf) demonstrated via simulation how atomized individuals evolve to be grouped in similar clusters or countries, a model of culture.^[[Axelrod, Robert. 1997. "The Dissemination of Culture." _Journal of Conflict Resolution_ 41(2): 203–26.](http://www-personal.umich.edu/~axe/research/Dissemination.pdf)]
* [Chen and Rodden (2013)](http://www-personal.umich.edu/~jowei/florida.pdf) argued in a 2013 article that the vote-seat inequality in U.S. elections that is often attributed to intentional partisan gerrymandering can actually attributed to simply the reality of "human geography" -- Democratic voters tend to be concentrated in smaller area. Put another way, no feasible form of gerrymandering could spread out Democratic voters in such a way to equalize their vote-seat translation effectiveness. After demonstrating the empirical pattern of human geography, they advance their key claim by simulating thousands of redistricting plans and record the vote-seat ratio.^[[Chen, Jowei, and Jonathan Rodden.  "Unintentional Gerrymandering: Political Geography and Electoral Bias in Legislatures. _Quarterly Journal of Political Science_, 8:239-269"](http://www-personal.umich.edu/~jowei/florida.pdf)]
* [Gary King, James Honaker, and multiple other authors](https://gking.harvard.edu/files/abs/evil-abs.shtml) propose a way to analyze missing data with a method of multiple imputation, which uses a lot of simulation from a researcher's observed dataset.^[[King, Gary, et al. "Analyzing Incomplete Political Science Data: An Alternative Algorithm for Multiple Imputation". _American Political Science Review_, 95: 49-69.](https://gking.harvard.edu/files/abs/evil-abs.shtml)]  (Software: Amelia^[[James Honaker, Gary King, Matthew Blackwell (2011). Amelia II: A Program for Missing Data. Journal of
  Statistical Software, 45(7), 1-47.](http://www.jstatsoft.org/v45/i07/)])

Statistical methods also incorporate simulation: 
  
* The bootstrap: a statistical method for estimating uncertainty around some parameter by re-sampling observations. 
* Bagging: a method for improving machine learning predictions by re-sampling observations, storing the estimate across many re-samples, and averaging these estimates to form the final estimate. A variance reduction technique. 
* Statistical reasoning: if you are trying to understand a quantitative problem, a wonderful first-step to understand the problem better is to simulate it! The analytical solution is often very hard (or impossible), but the simulation is often much easier :-) 


### Where are we? Where are we headed? {-}

Up till now, you should have covered:

* `R` basics
* Visualization
* Matrices and vectors
* Functions, objects, loops
* Joining real data


In this module, we will start to work with generating data within R, from thin air, as it were. Doing simulation also strengthens your understanding of Probability (Section \@ref{probability}).


### Check your Understanding {-}

* What does the `sample()` function do?
* What does `runif()` stand for?
* What is a `seed`?
* What is a Monte Carlo?


Check if you have an idea of how you might code the following tasks:

* Simulate 100 rolls of a die
* Simulate one random ordering of 25 numbers
* Simulate 100 values of white noise (uniform random variables)
* Generate a "bootstrap" sample of an existing dataset

We're going to learn about this today!


## Pick a sample, any sample

## The `sample()` function

The core functions for coding up stochastic data revolves around several key functions, so we will simply review them here.  

Suppose you have a vector of values `x` and from it you want to randomly sample a sample of length `size`. For this, use the `sample` function
```{r}
sample(x = 1:10, size = 5)
```

There are two subtypes of sampling -- with and without replacement.

1. Sampling without replacement (`replace = FALSE`) means once an element of `x` is chosen, it will not be considered again:
```{r}
sample(x = 1:10, size = 10, replace = FALSE) ## no number appears more than once
```

2. Sampling with replacement (`replace = TRUE`) means that even if an element of `x` is chosen, it is put back in the pool and may be chosen again. 
```{r}
sample(x = 1:10, size = 10, replace = TRUE) ## any number can appear more than once
```

It follows then that you cannot sample without replacement a sample that is larger than the pool.
```{r, error = TRUE}
sample(x = 1:10, size = 10, replace = FALSE)
```


So far, every element in `x` has had an equal probability of being chosen. In some application, we want a sampling scheme where some elements are more likely to be chosen than others. The argument `prob` handles this.

For example, this simulates 20 fair coin tosses (each outcome is equally likely to happen)
```{r}
sample(c("Head", "Tail"), size = 20, prob = c(.5, .5), replace = TRUE)
```


But this simulates 20 biased coin tosses, where say the probability of Tails is 4 times more likely than the number of Heads
```{r}
sample(c("Head", "Tail"), size = 20, prob = c(0.2, 0.8), replace = TRUE)
```

### Sampling rows from a dataframe

In tidyverse, there is a convenience function to sample rows randomly: `slice_sample()`

For example, load the dataset on cars, `mtcars`, which has 32 observations.

```{r}
mtcars
```
sample_n picks a user-specified number of rows from the dataset:

```{r}
mtcars[sample(1:ncol(mtcars), size=3, replace=F),]
slice_sample(mtcars, n=3)
```

Sometimes you want a X percent sample of your dataset. In this case use the `prop` argument to `slice_sample`

```{r}
mtcars %>% slice_sample(prop=.5)
```


A common task in statistics is to generate "bootstrap" samples of your data. This involves sampling a new dataset of equivalent size (`n` rows) but **with replacement** (so some observations are dropped; others might appear 2 or 3 times). You can do this with `slice_sample` as well

```{r}
mtcars %>% slice_sample(prop=1, replace=T)
```

As a side-note, these functions have very practical uses for any type of data analysis:

* Inspecting your dataset: using `head()` all the same time and looking over the first few rows might lead you to ignore any issues that end up in the bottom for whatever reason.
* Testing your analysis with a small sample: If running analyses on a dataset takes more than a handful of seconds, change your dataset upstream to a fraction of the size so the rest of the code runs in less than a second. Once verifying your analysis code runs, then re-do it with your full dataset (by simply removing the `sample_n` / `sample_frac` line of code in the beginning). While three seconds may not sound like much, they accumulate and eat up time.


## Random numbers from specific distributions

### `rbinom()` {-}
`rbinom` builds upon `sample` as a tool to help you answer the question -- what is the _total number of successes_ I would get if I  sampled a binary (Bernoulli) result from a test with `size` number of trials each, with a event-wise probability of `prob`. The first argument `n` asks me how many such numbers I want.

For example, I want to know how many Heads I would get if I flipped a fair coin 100 times. 
```{r}
rbinom(n = 100, size = 1, prob = 0.1)
```

Now imagine this I wanted to do this experiment 10 times, which would require I flip the coin 10 x 100 = 1000 times! Helpfully, we can do this in one line
```{r}
rbinom(n = 10, size = 100, prob = 0.5)
```

### `runif()` {-}
`runif` also simulates a stochastic scheme where each event has equal probability of getting chosen like `sample`, but is a continuous rather than discrete system.  We will cover this more in the next math module.

The intuition to emphasize here is that one can generate potentially infinite amounts (size `n`) of noise that is a essentially random


```{r}
runif(n = 5, min=0, max=1)
```


### `rnorm()` {-}

`rnorm` is also a continuous distribution, but draws from a Normal distribution -- perhaps the most important distribution in statistics. It runs the same way as `runif`

```{r}
rnorm(n=5, mean=0, sd=2)
```

To better visualize the difference between the output of `runif` and `rnorm`, let's generate lots of each and plot a histogram.

```{r}
from_runif <- runif(n = 1000)
from_rnorm <- rnorm(n = 1000)

par(mfrow = c(1, 2)) ## base-R parameter for two plots at once
hist(from_runif)
hist(from_rnorm)
```


## r, p, and d

Each distribution can do more than generate random numbers (the prefix `r`). We can compute the cumulative probability by the function `pbinom()`, `punif()`, and `pnorm()`. Also the density -- the value of the PDF -- by `dbinom()`, `dunif()` and `dnorm()`.  

```{r}

random_normals <- rnorm(n=1000000, mean=0, sd=1)
hist(random_normals)

random_binom_2 <- qbinom(runif(n=1000000), size=10, prob=.4)
hist(random_binom_2)
```


## `set.seed()`

`R` doesn't have the ability to generate truly random numbers! Random numbers are actually very hard to generate. (Think: flipping a coin --> can be perfectly predicted if I know wind speed, the angle the coin is flipped, etc.). Some people use random noise in the atmosphere or random behavior in quantum systems to generate "truly" (?) random numbers. Conversely, R uses deterministic algorithms which take as an input a "seed" and which then perform a series of operations to generate a sequence of random-seeming numbers (that is, numbers whose sequence is sufficiently hard to predict).

Let's think about this another way. Sampling is a stochastic process, so every time you run `sample()` or `runif()` you are bound to get a different output (because different random seeds are used). This is intentional in some cases but you might want to avoid it in others. For example, you might want to diagnose a coding discrepancy by setting the random number generator to give the same number each time. To do this, use the function `set.seed()`.

In the function goes any number. When you run a sample function in the same command as a preceding `set.seed()`, the sampling function will always give you the same sequence of numbers. In a sense, the sampler is no longer random (in the sense of unpredictable to use; remember: it never was "truly" random in the first place)

```{r}
set.seed(02138)
runif(n = 10)
```

The random number generator should  give you the exact same sequence of numbers if you precede the function by the same seed, 

```{r}
set.seed(02138)
runif(n = 10)
```

## Calculating an expectation using Monte Carlo

We can use repeated, independent draws from a random process in order to approximate its mean (or variance, or higher-order moments). The intuition comes from the law of large numbers. As we obtain repeated independent and identically distributed realizations from a random variable and take their average, this average should get closer and closer to the true expected value. The longer we run the simulation, the better the approximation.

Crucially, we can do this for essentially any random process that we can repeatedly obtain independent draws from. All we need to do is be able to implement the process in code. This can often be easier than trying to calculate an expectation analytically.

Consider the problem of calculating the expectation of the maximum of two independent 20-sided die rolls. We could construct a function that took the maximum of two independent draws from a 1-20 vector. Doing this a large number of times and taking the average allows us to approximate the true expectation of this new random variable without having to do any math!

```{r}
radvantage <- function(sides=20, k=2){
  return(max(sample.int(sides, k, replace = T)))
}

# Take the mean of 100,000 rolls with "advantage"
set.seed(60639)
mean(map_vec(1:1e5, radvantage))
```

Next, consider the "birthday problem" or the probability that among $k$ randomly chosen people, at least two people will have the same birthday. Recall that the expected value of an indicator random variable is equal to the probability that the indicator is equal to $1$ (so we can calculate probabilities using Monte Carlo by getting draws from an indicator variable for the event of interest). 

We could construct a function that simulates one hypothetical set of $k$ people. Assume no leap years and uniformly distributed birthdays.

```{r}
birthday <- function(k){
  # For each person, sample a birthday
  bdays <- map_vec(1:k, function(x) sample.int(365, 1))
  # If any birthdays are duplicated, the length of unique(x) will be shorter than the length of x
  return(as.numeric(length(bdays) != length(unique(bdays))))
}

set.seed(60639)
# For a room of 23 people
mean(map_vec(1:1e4, function(x) birthday(23)))

# For a room of 90 people
mean(map_vec(1:1e4, function(x) birthday(60)))
```

Notably, it is still important that the expectation of the random variable **exists**. Consider the *Cauchy* distribution which has no finite mean. Suppose I take a large number of independent draws from this distribution and average them. I get wildly different results across different runs!

```{r}
set.seed(60639)
mean(rcauchy(1e5)) # Average of 100,000 draws from a Cauchy(0, 1)
mean(rcauchy(1e5)) # Another average of 100,000 draws from a Cauchy(0, 1) -- wildly different resutl!
```


## Exercise {-}

###  Baccarat {-}

Monte Carlo simulations are frequently used to analyze expected outcomes in casino games and calculate the "house edge" or the percent of a player's winnings that they can expect to lose on a particular bet. One casino game that is known to have a very **low** house edge is Baccarat. In this exercise, you will implement a function that simulates a game of Baccarat. Using this function, you will calculate the expected value for each of the three types of bets in the game. You will then simulate how these expected values might change if the casino set different payouts for some of the bets. 

First, a brief primer on baccarat. Baccarat is a table game in which two hands of playing cards -- the "player" hand and the "banker" hand -- are drawn and then compared against one another. The hand containing the highest score wins the round. The most common version played in modern casinos is called ["punto banco"](https://en.wikipedia.org/wiki/Baccarat) baccarat and essentially plays itself according to a fixed set of card drawing rules. The names "player" and "banker" are simply labels for each of the two hands. 

Before each round of play, bettors can place one of three bets: that the "player" hand will win, that the "banker" hand will win, and that the hands will "tie" and have the same score.

Then, the round begins. The game is played by drawing cards from a "shoe" of cards containing multiple decks of standard playing cards (typically six or eight). The shoe is shuffled to randomize the order. **Two** cards are dealt to the *player hand* and **two** cards are dealt to the *banker hand*  We will ignore common casino practice of "burning" some cards from the top of the deck to discourage card counting as this does not impact the simulation. 

A third card may be then dealt to the player hand depending on the value of that hand. A third card may then also be dealt to the banker hand depending also on the value of the banker hand and whether the player was given a third card. The rules for determining whether either hand receives a third card are below. After all cards are dealt, the value of the hands is calculated. 

Face cards and 10s are worth zero points, aces are worth one point, and all other cards are worth their value (2-9). The value of a hand is the ones digit of the sum of the point values of the cards in that hand. So a hand consisting of a 2 and a 5 would be worth 7 points, while a hand consisting of a 4, 5 and 3 would be worth 2 points ($4 + 5 + 3 = 12, 12 \text{ mod } 10 = 2$). In other words, hands are valued at their point sum modulo 10.

The hand with the highest value is declared the winner. If the hands are equally valued, the result is a "tie"

The rules for determining whether the player or banker hand receives a third card in a given round are somewhat complex: 

**First**, if **either** the player or the banker has a hand valued at 8 or 9, then no third cards are drawn, the round ends, and a winner is declared based on the value of the initial two-card hands.

**Second**, the game decides whether to give the "player hand" a third card. If the player's initial hand value is 5 or less, then they are given a third card. If the player's initial hand value is 6 or 7, they "stand"

**Third**, the game decides whether to give the "banker hand" a third card. If the player did not receive a third card, the banker acts by the same rule as the player (draw if 5 or less, stand if 6 or 7). If the player did receive a third card, then the decision to draw depends on both the value of the banker's current hand and the value of the **third card drawn** by the player.

- If the banker's hand is valued at 2 or less, they always draw a third card irrespective of what the player was dealt.
- If the banker's hand is valued at 3, they draw a third card unless the player's third card is an 8.
- If the banker's hand is valued at 4, they draw a third card if the player's third card is between 2 and 7 (inclusive)
- If the banker's hand is valued at 5, they draw a third card if the player's third card is between 4 and 7 (inclusive)
- If the banker's hand is valued at 6, they draw a third card only if the player's third card is a 6 or 7.
- If the banker's hand is valued at 7, they do not draw a card

The Wikipedia page for Baccarat has a nice table that summarizes the "hit/stand" decision for the banker.

After the cards are dealt and the winner determined for the round, bets are paid. 

- If the bettor bet "player" and the "player" hand wins, they keep the amount wagered and receive an equal amount from the casino (pays 1-to-1)
- If the bettor bet "banker" and the "banker" hand wins, they keep the amount wagered and receive 95% of their wager from the casino (pays 19-to-20).
- If the bettor bet either "player" or "banker" and the result is a tie, they keep the amount wagered but receive no additional money from the casino (push).
- If the bettor bet "tie" and the result is a tie, they keep the amount wagered and receive 8 times their wagered amount from the casino (pays 8-to-1)
- If the bettor bet "player" and the "banker" wins or if they bet "banker" and the "player" wins, they lose the amount wagered.
- If the bettor bet "tie" and either "banker" or "player" wins, they lose the amount wagered.

### Part 1 (-)

Implement, in code, a function that simulates a single round of play (ignore the betting process for now). Assume the casino is playing with a shoe of six decks of standard playing cards and that this shoe is refreshed after every round.

Below is an outline for the sorts of functions you should implement to break this problem down into smaller component parts along with some hints.

```{r}
# This function should take as input some number of decks and generate a shoe of cards in the form of a vector
# HINT: You don't need to store the actual cards, just their values
# HINT 2: Read the documentation for the 'rep' function
gen_deck <- function(){

}
```

```{r}
# This function should take as input some vector that represents a hand of cards and return its value
# HINT: In R, the modulo operator is %%
value_hand <- function(){

}
```

```{r}
# This function should take as input the banker's and the player's hands and return the outcome of the round
determine_winner <- function(){

}
```

```{r}
# This is your main function. It should return (at a minimum) the outcome of the round of play
# HINT: Write out each phase of the round in plain language in the comments. Then try to implement that phase in code.
# HINT 2: You'll likely need to use a lot of conditional statements to implement the drawing rules
play_round <- function(){
 
}
```

### Part 2 (-)

Now implement a function that takes as input a bettor's chosen outcome and their wager. The function should then play a round of baccarat and return the amount that the bettor receives from the casino. There are a few equally valid ways to implement "winnings." For the purposes of this problem, you should have the function return $0$ if the player loses their entire wager.

```{r}
# This function should take as input a choice and a wager. You can choose to have it wrap `play_round()` or have the call to `play_round()` passed as an argument. Either way, it should return a "payoff" based on the choice the bettor selects and the (random) outcome of the round.
standard_payoff <- function(){
 
}
```

### Part 3 (-)

Using a Monte Carlo simulation, calculate the expected return of a wager of 100 dollars on the "player" bet. In other words, if a bettor pays $100$ to bet on "player", what is the amount that they expect to win. Set your seed once to $60639$ and run it for 300,000 iterations (try running for fewer iterations as you're testing, but you'll need a decent number of iterations to get the desired numerical precision)

Calculate the "house edge" of the "player" bet. The house edge is the difference between a bettor's amount wagered and their expected return, divided by the amount wagered. In other words, it's the amount that a casino expects to keep of a player's bet.

$$\text{HouseEdge} = \frac{\text{AmountWagered} - \text{ExpectedReturn}}{\text{AmountWagered}}$$

```{r}
set.seed(60639) # Set the seed
# Simulation code here

```

### Part 4 (-)

Use a Monte Carlo simulation to calculate the expected return of a wager of 100 dollars on the "banker" bet. Calculate the "house edge" of the "banker" bet. Set the seed to $60640$ and run for 300,000 iterations.

```{r}
set.seed(60640) # Set the seed
# Simulation code here

```

Which bet has the lower house edge?

### Part 5 (-)

One odd feature of Baccarat is that the "banker" bets pay 19-to-20 rather than the 1-to-1 for player bets. Using a Monte Carlo simulation, show why casinos don't pay 1-to-1 on the banker (calculate the house edge). You'll need to write a new payoff function to handle the alternative payout structure. Set the seed to $60641$ and run for 300,000 iterations.

```{r}
set.seed(60641) # Set the seed
# Simulation code here

```

### Challenge problem (-)

A new version of Baccarat that has become popular in some areas called "EZ-Baccarat" purports to solve the problem you identified in Part 5 while still paying "banker" bets at 1-to-1. It does so by having any round where the banker wins with a hand of 3 cards that totals 7 points be a "push" (bettors keep their wagers but don't win any additional money) rather than a "win" for the banker hand.

Show, using a Monte Carlo simulation, how this payout structure retains the house edge on "banker" bets. You may have to modify your simulation functions to return more information about the outcome of the round in order to implement a new payoff function.

See the [state regulatory documents](https://oag.ca.gov/sites/all/files/agweb/pdfs/gambling/BGC_ez_baccarat.pdf) for more info about this version of the game.

```{r}
set.seed(60642) # Set the seed

```


